Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-08-11 Manuscript/manuscript.tex

\subsection{Electronic continuum correction for lipid bilayers}
The lack of electronic polarizability in the standard MD simulation
force fields has been considered higly relevant issue since the early days of
lipid bilayer simulations.
%Some pioneering simulation studies scaled
%the partial charges of atoms by one half to effectively include 
%polarizability \cite{jonsson86,egberts94}. Also, approaches that
%explicitly include polarizability has been introduced, but
%this has turned out to be practically complicated \cite{chowdhary13}.
In this work we circumvent the rather demanding explicit inclusion
of electronic polariization effects \cite{chowdhary13} by effectively including
electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}. 
However, the concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13} both for derivation and application.

@@

Exploring different scaling factor values, applied to headgroup, glycerol backbone
and carbonyl charges of Lipid14 model, we found out that ion binding and
related headgroup order parameter response became weaker with the scaling of partial
charges. The optimal behaviour of ion binding was observed with
the scaling factor $f_q = 0.8$. % is optimal for applying ECC on Lipid14 model. 
Interestingly, this scaling factor is in line with the estimate
given by ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}
combined with RESP calculations in vacuum and implicit solvent reported
in \cite{maciejewski14}. IPolQ charges are obtained as the average of
partial charges given by RESP calculation~\cite{RESP_paper}
in vacuum and in a solvent. Applying the scaling factor of 0.75 to
IPolQ charges calculated from the data in Ref.~\cite{maciejewski14},
gives similar partial charges to ones obtained by scaling Lipid14 charges
with a factor 0.8.
\todo{SAMULI: I am still not sure if this is correct and/or makes sense.}
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of headgroup, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,??}.
This increases solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters of the same atoms for which charges were scaled
with a factor of $f_\sigma = 0.89$, the area per molecule value was
againg in agreement with experimental value (see Table~\ref{tab:apls}).

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}.
More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
Order parameters for all C-H bonds in lipid molecules
can be accurately measured by using H$^2$ NMR or $^{13}$C NMR techniques.
The same order parameters can be calculated from MD simulations directly
from the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

The change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Atomic absorption spectra and $^2$H NMR data, combined with the
information about order parameter signs gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.

Recent work published by the NMRlipids project showed that the concept works
qualitatively also in simulations and can be used to compare ion binding
affinitity between simulations and experiments~\cite{catte16}.
However, only systems with lipid bilayers and aqueous cations were used in that
study. The headgroup response to bound charge then depends on both, the
binding affinity and the response of the headgroup on bound charge.
Thus, the headgroup response to bound charge has to be correctly
described in the model if the concept is used for quantitative studies of
binding affinity. Hence, we first quantify the response of head group order
parameters to the bound charge by using experimental data measured from mixtures of cationinc surfactants
and PC lipids~\cite{scherer89}. The amount of the bound charge is known excatly in
these systems, because essentially all surfactants with unit charge locate in
lipid bilayers. Thus, the headgroup order parameter response to bound charge can be first
quantified aginst experimental data before proceeding to binding affinity studies.
%Secondly we evaluate the response of the head group order parameters
%to varying salt concentrations in solution
%and estimate the amount of membrane bound charge.

The experimental data used to quantify cation binding reports
order parameters as a function of equilibrium cation concentration
in the bulk solvent \cite{akutsu81,altenbach84}. Recent work
by NMRlipids project estimated this concentration by using
the molar concentration in water
[salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M,\\
\todo{SAMULI: I call here concentration we used to call ''nominal'' as ''molar concentration in water''.
  I think that the ''nominal'' is not a good term for this. We need to decide how we will call this.} \\
which was accurate enough for the conclusions in that study \cite{catte16}.
Since we are now targeting a model with quantitatively accurate binding
affinity, we used larger simulation box dimensions that allow the
determination of bulk ion conceration. Equilibrium concentration
corresponding the experimental data in the units of $\mathrm{mol/l}$
can be then determined from the ion concentration plateau in the bulk s
olvent at the boundaries of the simulation box as
\begin{equation}
\label{eq:conc}
  C_{eq}=\frac{C_{plateau}}{0.602},
\end{equation}
where plateau concentration is the number density in the units of $\mathrm{nm}^{-3}$.
\todo{SAMULI: Once we have to final results, we can probably say that
  the repeat distance is not far from the experimentally measured distance \cite{pabst07,uhrikova08}
  }

The binding affinities between different models were compared by calculating
the relative surface excess between water and ions $\Gamma_w^i$ \cite{chattorajBOOK}
\begin{equation}
  \Gamma_w^i=\Gamma_w^x-\Gamma_i^x\frac{c_w-c_w'}{c_i-c_i'},
\end{equation}
where $c_w$ and $c_i$ are concentrations in bulk water, 
$c_w'$ and $c_i'$ are concentrations inside bilayer, and
$\Gamma_w^x$ and $\Gamma_i^x$ are surface excess of water and
ions, respectively. Here consider the whole simulation box
as a interface region, thus the surface excesses
are $\Gamma_w^x=n_w/A$ and $\Gamma_i^x=n_i/A$, where
$A$ is the area of the simulation box, and $n_w$ and
$n_i$ are total number of water and ions in simulation box,
respectively. The concentrations of water and ions inside
bilayer can be expected to vanish, thus $c_w'=c_i'=0$.
Then we get the relative surface excess for lipid bilayer
simulations
\begin{equation}\label{surfexcess}
  \Gamma_w^i=\frac{n_w}{A}-\frac{n_i}{A}\frac{c_w}{c_i},
\end{equation}
where concentrations in bulk, $c_w$ and $c_i$, can be determined
from the simulation box boundary.

@@

Lipid bilayer structure without ions was validated against NMR experiments
by order parameters for C-H bonds and x-ray scattering experiments by using
form factors. The former validates the structures sampled by individual
lipid molecules in simulations with segmental resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}. Form factor was calculated from simulations
as 
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering lenght,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

@@

\subsubsection{Simulations with aqueous ions}
The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine
(POPC) bilayer and an aqueous salt solutions of varying concentrations.
Water molecules were described by OPC3 model \cite{Izadi16}.
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion model by Dang \cite{smith94,chang1999,dang2006}. 
Classical molecular dynamics simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}, 
and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}. 
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

\begin{table}[btp]
  \caption{Simulation parameters}
  \label{tbl:mdpar}
  \begin{tabular}{ll}
    simulation property & parameter   \\
    \hline
    time-step           & 2~fs         \\
    equilibration time  & 100~ns  \\
    simulation time     & 200~ns  \\
    temperature         & 313~K       \\
    thermostat          & v-rescale  \cite{bussi07}   \\
    barostat            & Parrinello-Rahman, semi-isotropic \cite{parrinello81} \\
    long-range electrostatics & PME  \cite{darden93}  \\
    cut-off scheme      & Verlet \cite{Pall13}      \\
    Coulomb and VdW cut-off & 1.0~nm \\
    constraints         & LINCS, only hydrogen atoms \cite{hess97} \\
    constraints for water & SETTLE  \cite{miyamoto92} \\
    \hline
  \end{tabular}
\end{table}

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
These systems were ran with original Lipid14 POPC model and with ECC-corrected model.
\todo{Describe the changes in surfactant model used with ECC-corrected lipids.}

@@

\subsection{Response of headgroup on bound charge}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.png}
  \caption{\label{OrderParameterCHANGESsurf}
    Headgroup order parameter changes and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$)
    in PC bilayer from simulations and experiments \cite{scherer89}.
  }
  \todo{Labels are missing.} \\
  \todo{Experimental data points are missing.}\\
  \todo{Order should be the same as in other figures, i.e, $\beta$ segment on top.}
\end{figure}

Headgroup order parameter response to the amount of bound charge can
be quantified by using ionic surfactants with a known valency~\cite{scherer89}.
All surfactants locate in lipid bilayers due to their amphiphilic nature,
thus the exact amount of bound charge in the membrane is known in these systems.
The headgroup order parameter changes as a function of bound charges can be then
explicitly determined from such systems.

Changes of headgroup order parameter as a function of added cationic surfactant
dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) from
experiments~\cite{scherer89} and simulations are shown in
Fig. \ref{OrderParameterCHANGESsurf}).
The head group order parameter response to the bound cation concentration
is approximately linear at least up to $\sim$0.3 mole fraction in experiments and
in both simulation models, original Lipid14 and ECClipids \todo{Yet to be shown.}.

@@

The headgroup P-N vector angle is shown in Fig.~\ref{??} as
function of bound cations.
As expected~\cite{seelig87}, the headgroup order parameter decrease due 
to bound cations is related to the orientration of headgroup more parallel
to membrane normal. The headgroup orientational effect is stronger
in Lipid14 parameters than in ECClipid parameters, which is in line
with smaller dipole moment and order parameter changes in ECClipid model.

@@

%After achieving accurate head group order parameter response with a cationic surfactant, 
%we advanced to quantifying the response to solutions with varying concentrations of salts, namely \ce{NaCl} and \ce{CaCl_2}.
The headgroup order parameter response to aqueous ionic concentration
can be used to measure ion binding affinity in experiments and simulations,
as discussed in section \ref{??} and in previous literature~\cite{seelig87,catte16}.
The headgroup order parameter response of PC lipids to NaCl and CaCl$_2$ concentrations
from experiments (DPPC \cite{akutsu81} and POPC \cite{altenbach84}) and different simulation
models are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
The simulation results of Lipid14 with Aqvist ions \cite{??} taken directly from
Ref. \citenum{catte16} are shown together with the results from Lipid14 simulation
with Dang ions \cite{??} and from ECClipid simulations with ECC-ions.

@@

Applying ECC correction also for lipids, introduced in section \ref{??},
brings the order parameter response to CaCl$_2$ concentration to a
good agreement with experiments, as shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
This is a significant improvement over all models introduced previously
which in general overstimate the head group order parameter response to CaCl$_2$
concentration \cite{catte16}. The good agreement with experiments indicates
that the ECC corrected lipid model has a sufficient accuracy for realistic
studies of ion binding details on lipid bilayers.

>>>>

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in the standard MD simulation
force fields has been considered higly relevant issue since the early days of
lipid bilayer simulations.
%Some pioneering simulation studies scaled
%the partial charges of atoms by one half to effectively include 
%polarizability \cite{jonsson86,egberts94}. Also, approaches that
%explicitly include polarizability has been introduced, but
%this has turned out to be practically complicated \cite{chowdhary13}.
In this work we circumvent the rather demanding explicit inclusion
of electronic polariization effects \cite{chowdhary13} by effectively including
electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}. 
However, the concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13} both for derivation and application.

@@

Exploring different scaling factor values, applied to headgroup, glycerol backbone
and carbonyl charges of Lipid14 model, we found out that ion binding and
related headgroup order parameter response became weaker with the scaling of partial
charges. The optimal behaviour of ion binding was observed with
the scaling factor $f_q = 0.8$. % is optimal for applying ECC on Lipid14 model. 
Interestingly, this scaling factor is in line with the estimate
given by ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}
combined with RESP calculations in vacuum and implicit solvent reported
in \cite{maciejewski14}. IPolQ charges are obtained as the average of
partial charges given by RESP calculation~\cite{RESP_paper}
in vacuum and in a solvent. Applying the scaling factor of 0.75 to
IPolQ charges calculated from the data in Ref.~\cite{maciejewski14},
gives similar partial charges to ones obtained by scaling Lipid14 charges
with a factor 0.8.
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of headgroup, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,??}.
This increases solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters with a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
again in agreement with experimental value (see Table~\ref{tab:apls}).

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}.
More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
Order parameters for $\alpha$ and $\beta$ segments in headgroup, as well as for
all C-H bonds in lipid molecules,
can be accurately measured by using H$^2$ NMR or $^{13}$C NMR techniques.
The experimental order parameters can be compared with the the ones
calculated from MD simulations by using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

According to the electrometer concept,
the change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Atomic absorption spectra and $^2$H NMR data, combined with the
information about order parameter signs gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
Recent work published by the NMRlipids project showed that the concept works
qualitatively also in simulations and can be used to compare ion binding
affinitity between simulations and experiments~\cite{catte16}.

The experimental data used to quantify cation binding reports
order parameters as a function of equilibrium cation concentration
in the bulk solvent \cite{akutsu81,altenbach84}. Recent work
by NMRlipids project estimated this concentration by using
the relative molar concentration in water
[salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M,
which was accurate enough for the conclusions in that study \cite{catte16}.
Since we are now targeting a model with quantitatively accurate binding
affinity, we used larger simulation box dimensions that allow the
determination of bulk ion conceration. Equilibrium concentration
corresponding the experimental data in the units of $\mathrm{mol/l}$
can be then determined from the ion concentration plateau in the bulk s
olvent at the boundaries of the simulation box as
\begin{equation}
\label{eq:conc}
  C_{eq}=\frac{C_{plateau}}{0.602},
\end{equation}
where plateau concentration is the number density in the units of $\mathrm{nm}^{-3}$.
\todo{SAMULI: Once we have to final results, we can probably say that
  the repeat distance is not far from the experimentally measured distance \cite{pabst07,uhrikova08}
  }

The ion binding affinities between different models were compared by calculating
the relative surface excess between water and ions $\Gamma_w^i$ \cite{chattorajBOOK}
\begin{equation}
  \Gamma_w^i=\Gamma_w^x-\Gamma_i^x\frac{c_w-c_w'}{c_i-c_i'},
\end{equation}
where $c_w$ and $c_i$ are concentrations in bulk water, 
$c_w'$ and $c_i'$ are concentrations inside bilayer, and
$\Gamma_w^x$ and $\Gamma_i^x$ are surface excess of water and
ions, respectively. The whole simulation box is considered here
as interfacial region, thus $\Gamma_w^x=n_w/A$ and $\Gamma_i^x=n_i/A$,
where $A$ is the area of the simulation box, and $n_w$ and
$n_i$ are total number of water and ions in simulation box,
respectively. The concentrations of water and ions inside
bilayer can be expected to vanish, thus $c_w'=c_i'=0$.
Then we get the relative surface excess for lipid bilayer
simulations
\begin{equation}\label{surfexcess}
  \Gamma_w^i=\frac{n_w}{A}-\frac{n_i}{A}\frac{c_w}{c_i},
\end{equation}
where concentrations in bulk, $c_w$ and $c_i$, can be determined
from the simulation box boundary.

@@

Lipid bilayer structure without ions was validated against NMR experiments
by order parameters for C-H bonds and x-ray scattering experiments by using
form factors. The former validates the structures sampled by individual
lipid molecules in simulations with segmental resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}.

Order parameters were calculated from simulations for all C-H bonds
by using Eq. \ref{OP} and form factors as
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering lenght,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

@@

\subsubsection{Simulations with aqueous ions}
The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine
(POPC) bilayer and an aqueous salt solutions of varying concentrations.
Water molecules were described by OPC3 model \cite{Izadi16}
\todo{Explain why were OPC3 chosen}.
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion model by Dang \cite{smith94,chang1999,dang2006}. 
Classical molecular dynamics simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}, 
and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}. 
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}. \\
\todo{System sizes has to reported somewhere}

\begin{table}[btp]
  \caption{Simulation parameters}
  \label{tbl:mdpar}
  \begin{tabular}{ll}
    simulation property & parameter   \\
    \hline
    time-step           & 2~fs         \\
    equilibration time  & 100~ns  \\
    simulation time     & 200~ns  \\
    temperature         & 313~K       \\
    thermostat          & v-rescale  \cite{bussi07}   \\
    barostat            & Parrinello-Rahman, semi-isotropic \cite{parrinello81} \\
    long-range electrostatics & PME  \cite{darden93}  \\
    cut-off scheme      & Verlet \cite{Pall13}      \\
    Coulomb and VdW cut-off & 1.0~nm \\
    constraints         & LINCS, only hydrogen atoms \cite{hess97} \\
    constraints for water & SETTLE  \cite{miyamoto92} \\
    \hline
  \end{tabular}
  \todo{This could be moved to SI. Only simulation lengths needs to be mentioned in the main paper.}
\end{table}

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35 or 50 surfactants per leaflet.
These systems were first simulated 200~ns by using Lipid14 model.
First 20~ns was omitted from the analysis.

The same systems where also simulated with ECC-corrected lipi model...
\todo{Describe the changes in surfactant model used with ECC-corrected lipids.}

@@

\subsection{Response of POPC headgroup on bound charge}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.eps}
  \caption{\label{OrderParameterCHANGESsurf}
    Headgroup order parameter changes and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$)
    in PC bilayer from simulations and experiments \cite{scherer89}.
  }
  \todo{Labels are missing.} \\
  \todo{Order should be the same as in other figures, i.e, $\beta$ segment on top.} \\
  \todo{I would put x-axis from 0 to 51 and maybe zoom y-axis little bit as well.}
\end{figure}

The headgroup response to bound charge depends on both, the
binding affinity and the response of the headgroup on bound charge.
Thus, the headgroup response to bound charge has to be correctly
described in the model if the concept is used for quantitative studies of
binding affinity. Hence, we first quantify the response of head group order
parameters to the bound charge by using experimental data measured from
mixtures of cationinc surfactants and PC lipids~\cite{scherer89}.
The amount of the bound charge is known excatly in these systems,
because essentially all surfactants with unit charge locate in lipid bilayers.
%Thus, the headgroup order parameter response to bound charge can be first
%quantified aginst experimental data before proceeding to binding affinity studies.
%Headgroup order parameter response to the amount of bound charge can
%be quantified by using ionic surfactants with a known valency~\cite{scherer89}.
%All surfactants locate in lipid bilayers due to their amphiphilic nature,
%thus the exact amount of bound charge in the membrane is known in these systems.
%The headgroup order parameter changes as a function of bound charges can be then
%explicitly determined from such systems.

Changes of headgroup order parameter as a function of added cationic surfactant
dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) from
experiments~\cite{scherer89} and simulations are shown in
Fig. \ref{OrderParameterCHANGESsurf}).
As expected from Eq. \ref{OPchangeEQ}, the head group order parameter response to the bound cation concentration
is approximately linear at least up to $\sim$0.3 mole fraction in experiments and
in both simulation models, original Lipid14 and ECClipids \todo{Yet to be shown.}. 
\todo{We might want to calculate the slopes from simulations.}

@@

Also the angle of the headgroup P-N vector with respect to membrane normal
is shown in Fig.~\ref{OrderParameterCHANGESsurf} as a function of bound cations.
As expected~\cite{seelig87}, the headgroup orients more parallel with membrane
normal with increasing amount of bound cations. The effect is stronger
in Lipid14 parameters than in ECClipid parameters, which is in line
with the reduced dipole moment and the lower sensitivity to bound charges
in the ECClipid model.
\todo{It might be a good idea to establish a relation between order parameter change
and P-N vector angle by using ECClipids model.}

@@

%After achieving accurate head group order parameter response with a cationic surfactant, 
%we advanced to quantifying the response to solutions with varying concentrations of salts, namely \ce{NaCl} and \ce{CaCl_2}.
The headgroup order parameter response to aqueous ionic concentration
can be used to measure ion binding affinity in experiments and simulations,
as discussed in section \ref{section:electrometer} and in previous literature~\cite{seelig87,catte16}.
The headgroup order parameter response of PC lipids to NaCl and CaCl$_2$ concentrations
from experiments (DPPC \cite{akutsu81} and POPC \cite{altenbach84}) and different simulation
models are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
The simulation results of Lipid14 with Aqvist ions \cite{??} (from Ref. \citenum{catte16})
are shown together with the results from Lipid14 simulation
with Dang ions \cite{??} and from ECClipid simulations with ECC-ions.

@@

Applying ECC correction also for lipids, introduced in section \ref{section:ecc},
brings the order parameter response to CaCl$_2$ concentration to a
good agreement with experiments, as shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
This is a significant improvement over previously introduced models,
which all overestimate the head group order parameter response to CaCl$_2$
concentration \cite{catte16}. The good agreement with experiments indicates
that the ECC corrected lipid model has a sufficient accuracy for realistic
studies of ion binding details on lipid bilayers.
